---
output: fost::default_manuscript
---

```{r setup, message = FALSE, echo = FALSE}
devtools::load_all()
set_options()
library(pls)
library(dplyr)
library(ggplot2)
library(patchwork)
```

Total characters: 2300 + 400 / 3200 (including authors and title, excluding spaces; +400
characters for figure)

Title characters: 106 / 150

**Leg length, a marker of early childhood conditions, associates with specific clusters of serum fatty acids**

Luke W. Johnston, Zhen Liu, Ravi Retnakaran, Stewart B. Harris, 
Richard P. Bazinet, Anthony J. Hanley

**Background and aims**: 
Adverse early childhood conditions have been associated with greater risk for
adult chronic diseases such as type 2 diabetes (T2DM) and cardiovascular disease.
However, the specific mechanism of action is not well elucidated. Adult leg
length is an established biomarker of early childhood conditions. We aimed to
explore distinct clusters of a broad spectrum of serum fatty acids (FA) by
height and leg length.

**Materials and Methods**: 
Canadian adults (n=453) at risk for T2DM from the Prospective Metabolism and
Islet Cell Evaluation (PROMISE) cohort had detailed personal and metabolic
data collected, including the measurement of FA and stature.
The concentrations of 22 FA in the cholesteryl ester (CE), phospholipid
(PL), triacylglycerol (TG), and non-esterified (NE) fractions were quantified.
Height and sitting height were measured, which were used to compute leg to
height ratio (LHR). To identify clusters in the FA profile, we used the supervised
dimensionality reduction method partial least squares (PLS) with the stature
components as the constraining variables and the FA as the predictor variables.
Separate models were analyzed for height and LHR.

```{r pls_models, echo=FALSE}
pls_data <- project_data %>% 
    filter(VN == 0) %>% 
    select_at(c(fa_conc_vars, outcome_vars)) %>% 
    na.omit()

# Height
pls_results <-
    pls_model(.data = pls_data,
              .xvar = fa_conc_vars,
              .yvar = outcome_vars[1],
              ncomp = 2) 
expl_ht <- pls::explvar(pls_results) %>% 
    sum() %>% 
    round(2)

fig_ht_load <- pls_results %>% 
    plot_loadings(load_range = c(-0.2, 0.2)) +
    coord_cartesian(xlim = c(-0.3, 0.35)) +
    ggtitle("Height")
fig_ht_corr <- pls_results %>% 
    plot_corr_comps() +
    ggtitle("Height")

# LHR
pls_results <-
    pls_model(.data = pls_data,
              .xvar = fa_conc_vars,
              .yvar = outcome_vars[4],
              ncomp = 2) 
expl_lhr <- pls::explvar(pls_results) %>% 
    sum() %>% 
    round(2)

fig_lhr_load <- pls_results %>% 
    plot_loadings(load_range = c(-0.2, 0.2)) +
    coord_cartesian(xlim = c(-0.3, 0.35)) +
    ggtitle("LHR")
    
fig_lhr_corr <- pls_results %>% 
    plot_corr_comps() +
    ggtitle("LHR")
```

```{r inline, echo=FALSE}
yr0 <- filter(project_data, VN == 0)
female <- calc_discr_npct(yr0$Sex, "Female")
european <- calc_discr_npct(yr0$Ethnicity, "European")
fa <- calc_fa_percent(yr0)
top_four_fa <- fa$c[1:4] %>% 
    stringr::str_c(collapse = ", ")
age <- aide::ave_sd(yr0$Age)
height <- aide::ave_sd(yr0$Height)
lhr <- aide::ave_sd(yr0$LegHtRatio, digits = 2)
```

**Results**:
The participants were mostly female (`r female$pct`) and of European-ancestry 
(`r european$pct`). Mean (SD) age was `r age` years, height was `r height` cm, and LHR was
`r lhr`. The four FA with the largest proportion in serum were `r top_four_fa`.
For each PLS model, we extracted the first two components (C1 and C2; Figure).
Higher LHR tended to load with a higher C1 cluster (i.e. more 20:5n-3 and 22:6n-3 in
multiple lipid fractions) and loaded with a lower C2 cluster (i.e. less 14:1n-7,
14:0, 16:0, 16:1n-7, 18:0 in primarily the TG and PL fractions). There were no
well defined specific cluster of FA in C1 for height, which may reflect that higher
height correlates positively with total FA concentration. Height tended to load
with a higher C2 cluster (e.g. 20 or more carbon long FA in multiple fractions).

**Conclusion**:
We found that shorter adult leg length had a distinct lipid profile compared to
shorter height, reflecting more omega-3 long chain FA and less of the 14 and 16
chain FA. Previous research has shown that these 14-16 chain FA associate with
greater de novo lipogenesis and exert lipotoxic effects. Our results suggest
that early childhood conditions, as reflected in adult leg length, may lead to
changes in lipid production and usage.

```{r fig_loadings, fig.height=5, echo=FALSE, dpi=600, dev="jpeg"}
(fig_ht_load + fig_lhr_load) +
        # (fig_ht_corr | fig_lhr_corr) +
    plot_layout(nrow = 2) +
    plot_annotation(caption = stringr::str_wrap("PLS loadings of C1 and C2 by height (A) and LHR (B). Higher loading values indicate a stronger contribution of the FA to the component. Interpretation is that a higher loading score correlates with a higher height or LHR, and that higher FA corresponds to its loading value. For example, in Figure B, one unit increase in PL 22:6 n-3 contributes to the C1 score by 0.269.", width = 100),
                    tag_levels = "A")
```

